Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-07-25 draft_lm1.tex
first commit of LM reviews; description of the simulation study

% Template for PLoS
% Version 2.0 July 2014
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
\documentclass[10pt]{article}
\usepackage{todonotes}
\usepackage{subfigure}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage[numbers]{natbib}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\usepackage{indentfirst}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{naturemag}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Our defs 
\def \rr {$R_{t}\:$}

% Leave date blank
\date{}

\pagestyle{myheadings}

@@

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\citep{wilson} 
and~\citep{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, W) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\citep{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 
- 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the \textbf{Appendix}.

@@

Figure~\ref{fig:rtseries}  shows the $R_t$ series, according to 
Equation~\ref{eq:Rtestimate}~\citep{nishiura} along with the confidence bands 
derived 
in this paper. 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1.
Due to the intrinsic variability of the \rr series, the examination of its 
credible intervals is essential to identify periods of sustained transmission.
The wider intervals between epidemics are due to the scarcity of cases during 
these periods.
The method to quantify uncertainty proposed here provides more conservative 
credibility intervals, and therefore offers protection against false alarms.
\begin{center}
\begin{figure}[!h]
 \centering
 \includegraphics[width=16cm]{./plots/rt_series.png}
 \caption{{\bf Estimated time-series for $R_t$, along with 95\% credible 
intervals.} Top panel show reported cases from which $R_t$ is estimated.
As expected, uncertainty about \rr is greater when the case counts are low, for 
instance in the period $2003$--$2006$, which represented a big hiatus between 
major epidemics.
The intrinsic variability of \rr can be used to inform the time-varying force 
of infection, since it reflects variation in the vector population and other 
environmental factors such as temperature and seasonal variation.}
\label{fig:rtseries}
\end{figure}
\end{center}

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
 Underreporting of cases, which is known to exist but of which exact figures 
cannot be determined, will lead to underestimation $S_0$. However, The attack 
ratio shall remain unbiased as the underreporting affects both the numerator and 
the denominator of equation \ref{eq:AR2}.
It is interesting to notice that the larger epidemics, in terms of peak size 
are not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population. Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$. Hon\'orio et 
al.\citep{honorio2009spatial} conducted a serological survey in three separate 
localities within the city, right before the 2008 epidemic, the authors report 
seroprevalences varying from 56-77.4\% which is compatible with our prediction 
of 87.5\% ($1-S_{0,2007}$) for the entire city, considering that we 
underestimate $S_0$ due to the underreporting of cases.

@@

\section*{Appendix}
\subsection*{A remark on prior distributions and tail behaviour of the 
distribution of $R_t$}
\label{sec:tails}
There are a number of approaches to deriving the distribution of \rr.
Alternatively to the approach described in the main text~\citep{mantel}, one 
could use the conditional distribution of \rr on 
$Y_{t+1}$ and $Y_t$ as defined in equation A7 of Nishiura et 
al.~\citep{nishiura}:
\begin{equation}
\label{seq:unorm}
f_{R}(R_{t}) = (Y_tR_{t})^{Y_{t+1}} e^{-Y_tR_{t}}
\end{equation}
Noticing the kernel of (\ref{seq:unorm}) is that of a gamma distribution with 
$a_2 = Y_{t+1}+1$ and $b_2 = Y_t$, we obtain a proper density from which to 
construct $c_{\alpha}(R_t)$, simply by computing the appropriate quantiles of 
said distribution.
 This density is
\begin{equation}
\label{seq:densityNishiura}
f_N(R_t| a_2, b_2) =  \frac{b_2^{a_2}}{\Gamma(a_2)} R_t^{a_2-1} e^{-b_2 R_t}
\end{equation}

In order to decide which approach to take, it may be of use analysing the 
tail behaviour of the derived distributions for \rr. 
Consider the case of using a flat $Uniform(0, 1)$ prior for $\theta_t$.
With $a_0 = b_0 = 1$, $a_1 = a_2$ and $b_1 = b_2 + 1$.
The beta prime (inverse beta distribution) will have heavier tails compared to 
the conditional distribution proposed by~\citep{nishiura}, thus providing more 
conservative confidence/credibility intervals.
To see that one needs simply take the ratio of the Beta prime and Gamma 
(unnormalized) densities and evaluate the limit as $R_t$ goes to infinity:
\begin{equation}
 \label{seq:densityratio}
 \lim_{R_t\to\infty}\frac{f_P(R_t| a_1, b_1)}{f_N(R_t| a_2, b_2)} =  
\lim_{R_t\to\infty}\frac{e^{Y_{t}R_t}}{(1 +R_t)^{Y_{t} + Y_{t +1}+2}} = \infty
\end{equation}
Finally, note that we deliberately construct $c_{\alpha}(R_{t})$ as a 
equal-tailed $100\alpha\%$ credible set, rather than a less conservative 
highest posterior density (HPD) interval.

@@

An advantage of the Bayesian approach is that one can devise prior 
distributions for $\theta_t$ taking advantage of the intuitive parametrization 
and flexibility of the beta family of distributions.
Prior elicitation can also be done for \rr and the hyper-parameters directly 
plugged into the prior for $\theta_t$. 
One can, for example, choose prior mean and variance for \rr and find $a_0$ 
and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{seq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and the coefficient of variation $c = 
\sqrt{v_0}/ m_0$ for $R_t$ \textit{a priori}, some less boring 
algebra gives:
\begin{align}
\label{seq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}

>>>>

% Template for PLoS
% Version 2.0 July 2014
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
\documentclass[10pt]{article}
\usepackage{todonotes}
\usepackage{subfigure}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage[numbers]{natbib}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\usepackage{indentfirst}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{naturemag}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Our defs 
\def \rr {$R_{t}\:$}
% Leave date blank
\date{}
\pagestyle{myheadings}

@@

Many authors have chosen to quantify the uncertainty about $\theta$ 
following orthodox approaches  (see for example~\citep{wilson} 
and~\citep{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, W) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$~\citep{dubey1970}.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 
- 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Note that this result holds only for $n = 1$.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior 
distributions for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in the~\textbf{Appendix}.

@@

Figure~\ref{fig:rtseries}  shows the $R_t$ series, according to 
Equation~\ref{eq:Rtestimate}~\citep{nishiura} along with the confidence bands 
derived 
in this paper. 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1.
Due to the intrinsic variability of the~\rr series, the examination of its 
credible intervals is essential to identify periods of sustained transmission.
The wider intervals between epidemics are due to the scarcity of cases during 
these periods. 
credibility intervals, and therefore offers protection against false alarms (see 
the section on tail behaviour in the \textbf{Appendix} for a detailed 
explanation).
\begin{center}
\begin{figure}[!h]
 \centering
 \includegraphics[width=16cm]{./plots/rt_series.png}
 \caption{{\bf Estimated time-series for $R_t$, along with 95\% credible 
intervals.} Top panel show reported cases from which $R_t$ is estimated.
As expected, uncertainty about \rr is greater when the case counts are low, for 
instance in the period $2003$--$2006$, which represented a big hiatus between 
major epidemics.
The intrinsic variability of \rr can be used to inform the time-varying force 
of infection, since it reflects variation in the vector population and other 
environmental factors such as temperature and seasonal variation.}
\label{fig:rtseries}
\end{figure}
\end{center}

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
 Underreporting of cases, which is known to exist but of which exact figures 
cannot be determined, will lead to underestimation $S_0$. However, The attack 
ratio shall remain unbiased as the underreporting affects both the numerator and 
the denominator of equation~\ref{eq:AR2}.
It is interesting to notice that the larger epidemics, in terms of peak size 
are not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population. Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$. Hon\'orio et 
al.\citep{honorio2009spatial} conducted a serological survey in three separate 
localities within the city, right before the 2008 epidemic, the authors report 
seroprevalences varying from 56-77.4\% which is compatible with our prediction 
of 87.5\% ($1-S_{0,2007}$) for the entire city, considering that we 
underestimate $S_0$ due to the underreporting of cases.

@@

\section*{Appendix}
\subsection*{A remark on prior distributions and tail behaviour of the 
distribution of $R_t$}
\label{sec:tails}
There are a number of approaches to deriving the distribution of~\rr.
Alternatively to the approach described in the main text~\citep{mantel}, one 
could use the conditional distribution of \rr on 
$Y_{t+1}$ and $Y_t$ as defined in equation A7 of Nishiura et 
al.~\citep{nishiura}:
\begin{equation}
\label{seq:unorm}
f_{R}(R_{t}) = (Y_tR_{t})^{Y_{t+1}} e^{-Y_tR_{t}}
\end{equation}
Noticing the kernel of (\ref{seq:unorm}) is that of a gamma distribution with 
$a_2 = Y_{t+1}+1$ and $b_2 = Y_t$, we obtain a proper density from which to 
construct $c_{\alpha}(R_t)$, simply by computing the appropriate quantiles of 
said distribution.
 This density is
\begin{equation}
\label{seq:densityNishiura}
f_N(R_t| a_2, b_2) =  \frac{b_2^{a_2}}{\Gamma(a_2)} R_t^{a_2-1} e^{-b_2 R_t}
\end{equation}

In order to decide which approach to take, it may be of use analysing the 
tail behaviour of the derived distributions for \rr. 
Consider the case of using a flat $\text{Uniform}(0, 1)$ prior for $\theta_t$.
With $a_0 = b_0 = 1$, $a_1 = a_2$ and $b_1 = b_2 + 1$.
The beta prime (inverse beta distribution) will have heavier tails compared to 
the conditional distribution proposed by~\citep{nishiura}, thus providing more 
conservative confidence/credibility intervals.
To see that one needs simply take the ratio of the Beta prime and Gamma 
(unnormalized) densities and evaluate the limit as $R_t$ goes to infinity:
\begin{equation}
 \label{seq:densityratio}
 \lim_{R_t\to\infty}\frac{f_P(R_t| a_1, b_1)}{f_N(R_t| a_2, b_2)} =  
\lim_{R_t\to\infty}\frac{e^{Y_{t}R_t}}{(1 +R_t)^{Y_{t} + Y_{t +1}+2}} = \infty
\end{equation}
Note also that we deliberately construct $c_{\alpha}(R_{t})$ as a 
equal-tailed $100\alpha\%$ credible set, rather than a less conservative 
highest posterior density (HPD) interval.
Finally, we performed a simple simulation study to assess the result in equation
~\ref{seq:densityratio}.
The simulations were carried out as follows:
\begin{enumerate}
 \item Create a two-dimensional grid of values for $\lambda_1$ and $\lambda_2$, 
with rate values from $2$ to $1000$;
\item For each point $(\lambda_1^{(j)}, \lambda_2^{(j)})$ in the grid,
\begin{itemize}
 \item Generate $1000$ realisations of the random variables $Y_{1}^{(ij)} \sim 
\text{Poisson}(\lambda_1^{(j)})$ and $Y_{2}^{(ij)} \sim 
\text{Poisson}(\lambda_2^{(j)})$, with $i= 1, 2, \ldots, 1000$  and compute 
$R_{t}^{(ij)} = Y_{2}^{(ij)}/
Y_{1}^{(ij)}$;
\item Compute the $\alpha\%$ credibility/confidence intervals using the method 
proposed here $(\beta)$ and using the distribution proposed by Nishiura 
(2010)~\cite{nishiura} ($\Gamma$), denoted by $c_{\beta}(R_{t}^{(ij)}; 
\alpha)$ and 
$c_{\Gamma}(R_{t}^{(ij)}; \alpha)$ respectively;
\item Determine whether
\begin{itemize}
 \item [(i)] $ c_{\Gamma}(R_{t}^{(ij)}; \alpha) \subseteq 
c_{\beta}(R_{t}^{(ij)}; \alpha)$;
\item [(ii)] $c_{\beta}(R_{t}^{(ij)}; \alpha) \subseteq 
c_{\Gamma}(R_{t}^{(ij)}; \alpha)$ or;
\item [(iii)] there is overlap
\end{itemize}
\item Calculate $Pr(R_{t}^{(ij)} > 1)$ using both distributions
\end{itemize}
\end{enumerate}
The results show that the credibility intervals obtained using the distribution 
proposed here were larger than (i.e. contained) those computed using the 
distribution in~\ref{seq:densityNishiura} in $90.69\%$ of the simulations 
(integrating over the grid), while the converse was observed in $1.14\%$ of the 
cases.
Moreover, our method yielded lower $Pr(R_{t} > 1)$  in $79.48\%$ of the 
simulations, indicating it is indeed more robust to false alarm.
An R script to perform the simulation study described above can be found 
at~\url{
https://github.com/fccoelho/paperLM1/blob/master/R/credibility_comparison.R}.

@@

An advantage of the Bayesian approach is that one can devise prior 
distributions for $\theta_t$ taking advantage of the intuitive parametrization 
and flexibility of the beta family of distributions.
Prior elicitation can also be done for~\rr and the hyper-parameters directly 
plugged into the prior for $\theta_t$. 
One can, for example, choose prior mean and variance for~\rr and find $a_0$ 
and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{seq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and the coefficient of variation $c = 
\sqrt{v_0}/ m_0$ for $R_t$ \textit{a priori}, some less boring 
algebra gives:
\begin{align}
\label{seq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}
